Reversible Regime Change: Climate-driven phytoplankton community shifts in the Cariaco Basin, Venezuela

### 2025-07-27 agujournaltemplate.tex
Double checked and reread - off to Ago again

\title{Climate-driven phytoplankton community variability in the Cariaco Basin, Venezuela}

@@

Phytoplankton communities are integral to oceanic biogeochemical cycles and are sensitive indicators of climate-driven environmental variability.
Long-term time series of phytoplankton can capture changes in climate and ecosystems, allowing us to understand the effects of environmental change on local communities. This study investigates changes in the phytoplankton community of the Cariaco Basin, Venezuela, from 1995 to 2017, using monthly water column observations integrated with climate indices and meteorological data. We considered the Atlantic Multidecadal Oscillation (AMO) index and the Multivariate El Niño Southern Oscillation index (MEI v.2), both of which have been linked to variations in wind that drive upwelling in the Cariaco Basin.
Cluster analysis identified two distinct community states associated with a shift from high to low upwelling conditions in 2004, followed by a return to a community and conditions similar to those before the shift in 2014. In contrast, community diversity declined between 1998 and 2009, independent of changes in the upwelling regime.
To quantify the effect of climate and bottom-up drivers on phytoplankton community changes, we employed a gradient forest analysis and found that the AMO index was the strongest predictor of change, closely followed by in situ variables related to upwelling (sea surface temperature and nitrate concentration) and the MEI v.2 index. Our study emphasizes the impact of long-term (decadal) climatic oscillations, such as the AMO, on the phytoplankton community of a tropical coastal ecosystem through modulation of the upwelling system and highlights the resilient nature of the tropical marine ecosystem under global change.

@@

% Topic Overview
Phytoplankton are microscopic algae that form the foundation of the ocean food web. Because they respond quickly to environmental changes, studying them can show how marine ecosystems are affected by climate.
% Paper Overview
This study looked at how phytoplankton communities in the Cariaco Basin, off the coast of Venezuela, changed from 1995 to 2017. We compared microscopy cell counts and monthly water samples with weather data and two major climate indicators: the Atlantic Multidecadal Oscillation (AMO) index and the Multivariate El Niño Southern Oscillation index (MEI v.2), both linked to local wind and currents.
% Findings Summary
The results showed a major shift in the phytoplankton community in 2004, linked to a decrease in upwelling, with conditions partially returning to the previous state by 2014. Interestingly, community diversity declined between 1998 and 2009, although upwelling had not yet changed. Further analysis showed that the AMO index, along with sea surface temperature, MEI v.2 and nitrate levels in the water column, were the strongest predictors of changes in the phytoplankton community.
% Key Takeaways
These findings contribute to a broader understanding of how climate variability shapes marine phytoplankton communities and underscore the importance of sustained long-term observations in detecting and interpreting ecological responses to climate change.

@@

% Phytoplankton is important
Phytoplankton provide the foundation for marine food webs and play crucial roles in biogeochemical processes, \cite{falkowski_biogeochemical_1998}. Their short generation times and high assemblage diversity allow phytoplankton communities to respond relatively quickly to environmental changes, making them suitable indicators of ecosystem state and dynamics \cite{alvarez-cobelas_what_1998, barton_anthropogenic_2016, di_cavalho_temporal_2023}. Long-term time series of biological observations are crucial for understanding how phytoplankton communities respond to short-term processes and long-term changes in the local physical environment and climate \cite{carstensen_need_2014, henson_observing_2016}. Such time series are rare, particularly in the tropics \cite{clarke_does_2017}.

@@

% PARAGRAPH THREE - CHANGES IN CARIACO
Since the inception of the time series in 1995, the ecosystem of the Cariaco Basin has undergone marked changes. \citeA{taylor_ecosystem_2012} analyzed the time series data up to 2010 and found decadal trends of increasing sea surface temperature (SST), coinciding with a decrease in upwelling and phytoplankton bloom intensities, which the authors attributed to global warming. During the initial phase of the time series (1996–2004), the Cariaco Basin underwent consistent and pronounced seasonal upwelling cycles, driven by strong trade winds during winter and spring (December – April) \cite{mullerkarger_annual_2001, astor_seasonal_2003}. After 2004, a significant reduction in the strength of trade winds during the upwelling period \cite{taylor_ecosystem_2012} caused a regime change in the physical and biogeochemical system. This strongly affected the phytoplankton community and the entire ecosystem. Microscopy cell counts indicated a drastic reduction of large phytoplankton, particularly diatoms, as confirmed, albeit not with the same high significance, by analysis of pigment data \cite{pinckney_phytoplankton_2015}. In general, a shift was observed in the phytoplankton community toward smaller cell sizes, along with increased abundances of coccolithophores, cryptophytes, and other phytoflagellates, groups commonly found in stratified water columns \cite{lorenzoni_characterization_2015}. An analysis of pigment data indicated a deepening of the chlorophyll maximum, an increase in pigment diversity, and a reduction in the seasonal variability of community composition \cite{pinckney_phytoplankton_2015}.

% PARAGRAPH FOUR - EXPLANATION OF CHANGES // OPEN QUESTIONS!!
However, the shift in phytoplankton community composition associated with increased temperature and reduced upwelling intensities from 2004 to 2013 did not appear to be a permanent change in regime. Data from 2013 to 2017, the end phase of the CARIACO time series, indicated a return to strong upwelling intensities and high numbers of diatoms and dinoflagellates \cite{muller-karger_scientific_2019}. The impact of large-scale climatic processes on the observed shifts in the phytoplankton community remains unclear. The Atlantic Multidecadal Oscillation (AMO) has emerged as an important factor influencing the climate system in the region and the ecosystems of the north Atlantic \cite{nye_ecosystem_2014}. The oscillations of the AMO index are strongly correlated with temperature indices in the Caribbean region \cite{stephenson_changes_2014}. The El Niño Southern Oscillation (ENSO) is teleconnected to the Caribbean Sea, given that during ENSO (indicated by a positive MEI v.2 index) sea surface temperatures increase and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}. However, using the time series available only until 2010, \citeA{taylor_ecosystem_2012} found no significant correlation between MEI v.2 and in situ observations, and did not test for correlations with the AMO index.

% What drives the changes??? Bottom-up/ Top-down ... CLIMATE 
\citeA{taylor_ecosystem_2012} cautioned that the observed changes cannot be clearly identified as "unidirectional trends driven by anthropogenic climate change or whether they simply reflect low-frequency natural cycles, such as those driving the Atlantic Multidecadal Oscillation". Previous analyses of phytoplankton community diversity in the Cariaco Basin did not include time series data beyond 2011, and no detailed investigations have yet been conducted on changes in diversity based on microscopy cell counts. Given these knowledge gaps, we aim to investigate 1) changes in the depth distribution of phytoplankton biomass and community diversity over the period from 1995 to 2017, 2) changes in the structure of the phytoplankton community data and potential connections with shifts observed in physical and biogeochemical variables, and 3) potential connections between long-term climatic oscillations (ENSO MEI v.2 and AMO) and changes in the phytoplankton community. Our analysis focuses exclusively on bottom-up drivers of community changes because these data are available throughout the entire time series and at a monthly resolution. Our study builds on and substantially expands the work of \citeA{taylor_ecosystem_2012} and \citeA{pinckney_phytoplankton_2015} with a more detailed analysis of biodiversity aspects using all data available from the CARIACO time series up to the final cruise in 2017. Understanding the regional dynamics of the Cariaco Basin is crucial for unveiling the ecological changes affecting this tropical marine ecosystem and the variables driving these changes.

@@

    For the identification of phytoplankton species by microscopy, \qty{500}{\milli\liter} seawater samples were collected at depths of 1, 7, 15, 25, 35, 55, 75, and 100 m in HDPE bottles and preserved with a \qty{5}{\%} (final concentration) formalin solution neutralized with sodium tetraborate. Species counts were performed at the Universidad de Oriente, Venezuela, following the Utermöhl method \cite{hasle1978inverted}, using \qty{100}{\milli\liter} sedimentation chambers with a 48-hour settling period. Phytoplankton identification and enumeration by microscopy were restricted to microphytoplankton (cells $>$ \qty{20}{\micro \meter}) \cite{mutshinda_environmental_2013}.
    %from Mutshinda et al. 2013
    Quality control of the microscopy data was performed by assembling the taxonomic identifications made across all cruises, removing observations that were not identified at the species or genus level (with the exception of nanoflagellate cell counts), correcting spelling variations, and grouping synonyms using taxonomic information from the World Register of Marine Species (WoRMS, \url{marinespecies.org}). A total of 482 phytoplankton species were observed over the course of the time series, which were grouped into 223 phytoplankton genera. The identified genera were further grouped into functional groups based on their taxonomic classification in WoRMS.

@@

\subsection{Data analysis}    
    In the analysis, we set out to find if there is a structure in the distribution and temporal evolution of diversity metrics and the community composition and to identify potential bottom-up drivers of this change. 
    To provide an overview of the available data and highlight interannual trends and changes in the time series data, we calculated annual anomalies for relevant variables as normalized z scores ($z = \frac{(x – \mu)}{\sigma}$, where $\mu$ is the mean across all available data points). Z scores from 1995, 2016, and 2017 were not shown because insufficient coverage (i.e. more than six missing monthly samples) in these years prevented us from producing reliable yearly estimates.

@@

\subsection{Interannual changes in the environment and ecosystem}
    % Figure 2
    Coordinated, interannual changes are observable in normalized yearly means of all environmental variables considered, the logarithmic transformation of chlorophyll \textit{a}, and phytoplankton cell counts per functional group (Figure \ref{fig:zscore}). The patterns emerging from wind speed, evaporation, precipitation, SST, and \qty{21}{\celsius} isotherm depth clearly show the years of positive anomalies of wind (blue) and negative anomalies for SST (red) and upwelling (red, i.e., shallower depth of the isotherm). These patterns are reflected in the mode of the AMO, where a positive anomaly correlates with weak upwelling regimes (see also Figure \ref{fig:sup:correlation} in Appendix). The yearly mean nutrient concentrations have a maximum positive anomaly at the beginning of the time series and are highly correlated. Annual mean anomalies of nitrate and phosphate are generally high during stronger upwelling, but remain relatively low between 2002 and 2014, with some variability in mean phosphate concentration. Silicate does not follow the pattern of nitrate or phosphate, but shows an elevated mean concentration between 2008 and 2013, which could be driven by increased surface runoff \cite{lorenzoni_characterization_2015} and reduced depletion due to reduced abundances of diatoms.

    Coordinated, interannual changes are observable in normalized yearly means of all environmental variables considered, the logarithmic transformation of chlorophyll \textit{a}, and phytoplankton cell counts per functional group (Figure \ref{fig:zscore}). Distinct patterns emerge in wind speed, evaporation, precipitation, SST, and the \qty{21}{\celsius} isotherm depth. Periods of positive wind anomalies (blue) correspond to negative anomalies in SST and upwelling (red, indicated by a shallower isotherm depth), reflecting stronger upwelling conditions. These patterns are consistent with the Atlantic Multidecadal Oscillation (AMO), where a positive AMO phase is associated with weaker upwelling regimes (see also Figure~\ref{fig:sup:correlation} in the Appendix). Yearly mean nutrient concentrations show strong positive anomalies at the beginning of the time series and are generally well correlated. Nitrate and phosphate anomalies tend to be higher during periods of stronger upwelling but remain relatively low between 2002 and 2014, with nitrate showing more interannual variability. Silicate anomalies, in contrast, do not follow the same pattern as nitrate and phosphate. Instead, elevated concentrations are observed between 2008 and 2013, potentially driven by increased surface runoff \cite{lorenzoni_characterization_2015} and reduced consumption due to lower diatom abundance.
    Chlorophyll \textit{a} and functional group cell counts display positive anomalies throughout most of the first half of the time series, followed by a significant reduction after 2004, suggesting a shift. This shift occurs during the period between 2003 and 2013, when the AMO index presents mostly positive anomalies, while upwelling-related variables indicate a weak upwelling regime (low wind, high SST, deep isotherm). Although data are sparse, nutrients and biological variables show a strong increase following 2013, coinciding with a return to negative anomalies in the AMO index and more upwelling-favorable conditions.
    The annual means of diversity metrics do not follow this trend, but are reduced between 1998 and 2008. Shannon diversity specifically appears elevated past 2006 and the general pattern that emerges is that mean diversity is lower in years, where mean chlorophyll \textit{a} is high. However, in 1996 and 1997 positive anomalies in cell counts, chlorophyll \textit{a} and diversity metrics coincide.

    \begin{figure}
    \noindent\includegraphics[width=\textwidth]{fig/Figure2_ZScores_v2.pdf}
    \caption{Time series showing annual z-scores of selected environmental, climatological, and ecological variables. Red indicates negative annual anomalies and blue indicates positive annual anomalies. Deeper shades indicate larger anomalies. Chlorophyll and functional group counts were log-transformed prior to normalisation. The years 1995 (n=2) and 2016 (n=5), 2017 (n=1) were removed due to insufficient monthly data to generate robust annual anomalies. Additionally we marked years with two or three and four or more missing monthly values.}
    \label{fig:zscore}
    \end{figure}

    % Figure 3
    Figure \ref{fig:divts} shows the depth distribution and dynamics of chlorophyll \textit{a} and community diversity indices, also in relation to rainy and upwelling seasons. The bulk of the chlorophyll \textit{a} ($\sim$ 80\,\%) was measured in the upper 50 meters of the water column, with approximately 15\,\% measured between 50 and 75 meters. Chlorophyll \textit{a} exhibited marked seasonality, with the highest biomass in the upper water column during the early upwelling season (December–February, Figure \ref{fig:divts}$b$). The yearly mean chlorophyll \textit{a} biomass showed the highest variability within the upper 25 meters. In some years, particularly in 2000, 2005, 2007, 2008, and 2013, the concentration in this layer was lower than in the 25–50 meter depth range, suggesting a deep chlorophyll maximum in those years (Figure \ref{fig:divts}$a$). The year 2013 was an outlier, particularly for the upper 25 meters, as missing monthly sampling from the upwelling season (January and February) skewed the value downwards. Nevertheless, the monthly measurements for the remainder of the year 2013 show a markedly lower cholorophyll (65\,\% of the mean over the entire time series for these months), which recovered in 2014. Chlorophyll \textit{a} biomass below 50 meters depth showed a relatively stable distribution throughout the year and across the time series. In fall, during the rainy season, mean chlorophyll \textit{a} biomass in the upper 25 meters was markedly reduced, while the 50–75 meter depth interval showed a slight increase, indicating a shift in the chlorophyll \textit{a} maximum to lower depths under weak upwelling conditions.

    The calculated genus diversity from the microscopy cell counts generally follows the dynamics of chlorophyll \textit{a} biomass, with maximum values observed during the upwelling seasons in winter. Genus richness is consistently the highest diversity index in the top \qty{25}{meters} depth interval, also showing the highest variability over the time series. Across the time series, we observe maximum richness at the start of the time series (1996-1997), with a subsequent reduction in diversity across all depths, with a recovery after 2011. The reduction in diversity following the start of the time series does not follow the pattern shown by chlorophyll \textit{a}, which remains relatively stable across the top \qty{100}{meters} until 2004 (see Figure \ref{fig:divts $a$}). Scaling genus richness by the cell count abundances, i.e. the Shannon index, shows a similar but less pronounced dynamic. Diversity remains high throughout 2000 and recovers more quickly but shows marked reductions in annual means, particularly in the upper 25 meters in 2006, 2009, and 2016. The seasonal signal is lost (see Figure \ref{fig:divts} $f$).

    Community evenness, expressed by the Pielou index, shows no seasonal signal and a very large variance within seasons and through depth (see Figure \ref{fig:divts} $h$). The inter-annual dynamics of evenness are remarkably stable across all depth intervals, clustered around an intermediate 0.5 value. There are marked reductions in all depths towards 2000 and for all depths above 75 meters in 2006, 2009 and 2016. Throughout 2003 to 2013 evenness is almost consistently highest in the deepest interval below 75 meters, where the community seems to have been more evenly distributed.

@@

\subsection{Clustering analysis yields contrasting ecological regimes}
    % Figure 4
    We undertook clustering analysis on annually aggregated cell counts from microscopy and on individual monthly measurements. Figure \ref{fig:clustering}$a$ shows the results of clustering the annually aggregated communities using a linear complete linkage algorithm based on the binary Jaccard matrix. Two distinct clusters emerge from this analysis, with the first cluster encompassing the years 1996-2004 and 2014-2016, and the second cluster including the intermediate years from 2005 to 2013. 
    Inter-annual species turnover is shown in Figure \ref{fig:clustering}$b$, it reveals an increasing trend in turnover from 1996 to 2004. Turnover subsequently decreases during the period underpinning cluster 2 and increases again in 2014, marking a return to cluster 1. This pattern is also reflected in the hierarchical clustering results, Figure \ref{fig:clustering}$a$, where cluster 2 shows a lower distance in the clustering tree between the years. We additionally extracted the three dominant genera for the second cluster and the early and late parts of the first cluster, respectively. The diatom \textit{Thalassiosira} dominated the early period and was abundant in all other periods. \textit{Pseudo-nitzschia} dominates cluster 2, with the haptophyte \textit{Emiliana} being the second most abundant genus. The later part of cluster 1 is dominated by \textit{Emiliana}.

@@

    % Figure 5 - NEW
    Figure \ref{fig:clustcomp} shows density distributions of environmental variables, chlorophyll \textit{a}, and diversity indices for the two clusters. Upwelling-related variables exhibit distinct patterns between the two clusters. For example, cluster 1 experiences more frequent high wind events, a shallower \qty{21}{\celsius} isotherm, lower temperatures, reduced evaporation, and higher salinity and nutrient concentrations, particularly for nitrate. The AMO index shows a pronounced separation between the two clusters, with cluster 1 exhibiting more negative anomalies, while cluster 2 shows a higher occurrence of positive anomalies. Although there are no significant differences in genus richness and the Shannon index between clusters, there is a slightly increased median in the Pielou index for cluster 2 (Table \ref{tab:ClustCompWilcox}). The density distribution reveals some differences. Cluster 1 shows higher variance for genus richness, while cluster 2 shows greater variance and a bimodal distribution for the Shannon and Pielou indices.

@@

    The in-situ measurements that best predict changes in the microphytoplankton community were SST, nitrate concentration, and salinity, in that order. Nitrate is a crucial nutrient for phytoplankton growth and has been identified as the growth-limiting nutrient during stratification periods in the Cariaco Basin \cite{muller-karger_scientific_2019}. For diatoms, we observe a strong threshold response above \qty{7}{\micro \molar}, indicating that intense upwelling conditions drive blooms, resulting in predictable changes in cell counts. Similarly, haptophytes and diatoms exhibit a threshold around \qtyrange{22}{24}{\celsius}, indicating thresholds at low surface temperatures driven by stronger upwelling.

>>>>

\title{Reversible Regime Change: Climate-driven phytoplankton community shifts in the Cariaco Basin, Venezuela}

@@

Phytoplankton communities are integral to oceanic biogeochemical cycles and are sensitive indicators of climate-driven environmental variability.
Long-term time series can capture this variability, allowing us to untangle the effects of environmental change on local communities. This study investigates changes in the phytoplankton community of the Cariaco Basin, Venezuela, from 1995 to 2017, using monthly water column observations integrated with climate indices and meteorological data. To investigate impacts of larger scale climate, we considered the Atlantic Multidecadal Oscillation (AMO) index and the Multivariate El Niño Southern Oscillation index (MEI v.2), both of which have been linked to variations in wind that drive upwelling in the Cariaco Basin.
Cluster analysis identified two distinct community states associated with a shift from high to low upwelling conditions in 2004, followed by a return to a community and conditions similar to those before the shift in 2014. In contrast, community diversity declined between 1998 and 2009, independent of changes in the upwelling regime.
To quantify the effect of climate and bottom-up drivers on phytoplankton community changes, we employed a gradient forest analysis and found that the AMO index was the strongest predictor of change, closely followed by in situ variables related to upwelling (sea surface temperature and nitrate concentration) and the MEI v.2. Our study emphasizes the impact of long-term climatic oscillations, such as the AMO, on the phytoplankton community of a tropical coastal ecosystem through modulation of the upwelling system and highlights the resilient nature of the tropical marine ecosystem under global change.

@@

% Topic Overview
Phytoplankton are microscopic algae that form the foundation of the ocean food web. Because they respond quickly to environmental changes, studying them can show how marine ecosystems are affected by climate.
% Paper Overview
This study looked at how phytoplankton communities in the Cariaco Basin, off the coast of Venezuela, changed from 1995 to 2017. We compared microscopy cell counts and monthly water samples with weather data and two major climate indicators: the Atlantic Multidecadal Oscillation (AMO) index and the Multivariate El Niño Southern Oscillation index (MEI v.2), both linked to local weather and currents.
% Findings Summary
The results showed a major shift in the phytoplankton community in 2004, linked to a decrease in upwelling, with conditions partially returning to the previous state by 2014. Interestingly, community diversity declined between 1998 and 2009, although upwelling had not yet changed. Further analysis showed that the AMO index, along with sea surface temperature, MEI v.2 and nitrate levels in the water column, were the strongest predictors of changes in the phytoplankton community.
% Key Takeaways
These findings contribute to a broader understanding of how climate variability shapes marine phytoplankton communities and underscore the importance of sustained long-term observations in detecting and interpreting ecological responses to climate change.

@@

% Phytoplankton is important
Phytoplankton provide the foundation for marine food webs and play crucial roles in biogeochemical processes \cite{falkowski_biogeochemical_1998}. Their short generation times and high assemblage diversity allow phytoplankton communities to respond relatively quickly to environmental changes, making them suitable indicators of ecosystem state and dynamics \cite{alvarez-cobelas_what_1998, barton_anthropogenic_2016, di_cavalho_temporal_2023}. Long-term time series of biological observations are crucial for understanding how phytoplankton communities respond to both short- and long-term changes in the physical environment and climate \cite{carstensen_need_2014, henson_observing_2016}. Such time series are rare, particularly in the tropics \cite{clarke_does_2017}.

@@

% PARAGRAPH THREE - CHANGES IN CARIACO
Since the inception of the time series in 1995, the ecosystem of the Cariaco Basin has undergone marked changes. \citeA{taylor_ecosystem_2012} analyzed the time series data up to 2010 and found decadal trends of increasing sea surface temperature (SST), coinciding with a decrease in upwelling and phytoplankton bloom intensities, which the authors attributed to effects of global warming. During the initial phase of the time series (1996–2004), the Cariaco Basin underwent consistent and pronounced seasonal upwelling cycles, driven by strong trade winds during winter and spring (December – April) \cite{mullerkarger_annual_2001, astor_seasonal_2003}. After 2004, a significant reduction in the strength of trade winds during the upwelling period \cite{taylor_ecosystem_2012} caused a regime change in the physical and biogeochemical system. This strongly affected the phytoplankton community and the entire ecosystem. Microscopy cell counts indicated a drastic reduction of large phytoplankton, particularly diatoms \cite{pinckney_phytoplankton_2015}. In general, a shift was observed in the phytoplankton community toward smaller cell sizes, along with increased abundances of coccolithophores, cryptophytes, and other phytoflagellates, groups commonly found in stratified water columns \cite{lorenzoni_characterization_2015}. An analysis of pigment data indicated a deepening of the chlorophyll maximum, an increase in pigment diversity, and a reduction in the seasonal variability of community composition \cite{pinckney_phytoplankton_2015}.

% PARAGRAPH FOUR - EXPLANATION OF CHANGES // OPEN QUESTIONS!!
However, the shift in phytoplankton community composition associated with increased temperature and reduced upwelling intensities from 2004 to 2013 did not appear to be a permanent change in regime. Data from 2013 to 2017, the final years of the CARIACO (CArbon Retention In A Colored Ocean) time series, indicated a return to strong upwelling intensities and high numbers of diatoms and dinoflagellates \cite{muller-karger_scientific_2019}. The impact of large-scale climatic processes on the observed shifts in the phytoplankton community remains unclear. The Atlantic Multidecadal Oscillation (AMO) has emerged as an important factor influencing the climate system in the region and the ecosystems of the north Atlantic \cite{nye_ecosystem_2014}. The oscillations of the AMO index are strongly correlated with temperature indices in the Caribbean Sea \cite{stephenson_changes_2014}. The El Niño Southern Oscillation (ENSO) is teleconnected to the Caribbean Sea, given that during ENSO (indicated by a positive MEI v.2 index) sea surface temperatures increase and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}. However, using the time series available only until 2010, \citeA{taylor_ecosystem_2012} found no significant correlation between MEI v.2 and in situ observations, and did not test for correlations with the AMO index.

% What drives the changes??? Bottom-up/ Top-down ... CLIMATE 
\citeA{taylor_ecosystem_2012} cautioned that the observed changes cannot be clearly identified as "unidirectional trends driven by anthropogenic climate change or whether they simply reflect low-frequency natural cycles, such as those driving the Atlantic Multidecadal Oscillation". Previous analyses of phytoplankton community diversity in the Cariaco Basin did not include time series data beyond 2011, and no detailed investigations have yet been conducted on changes in diversity based on microscopy cell counts. Given these knowledge gaps, we aim to investigate 1) changes in the depth distribution of phytoplankton biomass and community diversity over the period from 1995 to 2017, 2) changes in the structure of the phytoplankton community data and potential connections with shifts observed in physical and biogeochemical variables, and 3) potential connections between long-term climatic oscillations (ENSO MEI v.2 and AMO) and changes in the phytoplankton community. Our analysis focuses exclusively on bottom-up drivers of community changes because these data are available throughout the entire time series and at a monthly resolution. Our study builds on and substantially expands the work of \citeA{taylor_ecosystem_2012} and \citeA{pinckney_phytoplankton_2015} with a more detailed analysis of biodiversity aspects using all data available from the CARIACO time series up to the final cruise in 2017. In doing so, we hope to provide a more comprehensive assessment of long-term phytoplankton community dynamics in the Cariaco Basin, offering new insights into the ecological responses to environmental variability across decadal scales.

@@

    For the identification of phytoplankton species by microscopy, \qty{500}{\milli\liter} seawater samples were collected at depths of 1, 7, 15, 25, 35, 55, 75, and 100 m in HDPE bottles and preserved with a \qty{5}{\%} (final concentration) formalin solution neutralized with sodium tetraborate. Species counts were performed at the Universidad de Oriente, Venezuela, following the Utermöhl method \cite{hasle1978inverted}, using \qty{100}{\milli\liter} sedimentation chambers with a 48-hour settling period. Phytoplankton identification and enumeration by microscopy were restricted to microphytoplankton (cells $>$ \qty{20}{\micro \meter}) \cite{mutshinda_environmental_2013}.
    %from Mutshinda et al. 2013
    Quality control of the microscopy data was performed by assembling the taxonomic identifications made across all cruises, removing observations that were not identified at the species or genus level (with the exception of nanoflagellate cell counts that are shown in Figure \ref{fig:zscore}), correcting spelling variations, and grouping synonyms using taxonomic information from the World Register of Marine Species (WoRMS, \url{marinespecies.org}). A total of 482 phytoplankton species were observed over the course of the time series, which were grouped into 223 phytoplankton genera. The identified genera were further grouped into functional groups based on their taxonomic classification in WoRMS.

@@

\subsection{Data analysis}    
    In the analysis, we set out to find if there is a structure in the distribution and temporal evolution of diversity metrics and the community composition, as well as to identify potential bottom-up drivers of this change. 
    To provide an overview of the available data and highlight interannual trends and changes in the time series data, we calculated annual anomalies for relevant variables as normalized z scores ($z = \frac{(x – \mu)}{\sigma}$, where $\mu$ is the mean across all available data points). Z scores from 1995, 2016, and 2017 were not shown because insufficient coverage (i.e. more than six missing monthly samples) in these years prevented us from producing reliable yearly estimates.

@@

\subsection{Interannual changes in the environment and ecosystem}
    % Figure 2
    Coordinated, interannual changes are observable in normalized yearly means of all environmental variables, as well as log-transformed chlorophyll \textit{a} and phytoplankton cell counts per functional group (Figure \ref{fig:zscore}). Distinct patterns emerge in wind speed, evaporation, precipitation, SST, and the \qty{21}{\celsius} isotherm depth. Periods of positive wind anomalies (blue) correspond to negative anomalies in SST and upwelling (red, indicated by a shallower isotherm depth), reflecting stronger upwelling conditions. These patterns are consistent with the Atlantic Multidecadal Oscillation (AMO), where a positive AMO phase is correlated with weaker upwelling regimes (see also Figure~\ref{fig:sup:correlation} in the Appendix). Yearly mean nutrient concentrations show strong positive anomalies at the beginning of the time series and are generally well correlated. Nitrate and phosphate anomalies tended to be higher during periods of stronger upwelling but remained relatively low between 2002 and 2014, with nitrate showing more interannual variability. Silicate anomalies, in contrast, did not follow the same pattern as nitrate and phosphate. Instead, elevated concentrations were observed between 2008 and 2013, potentially driven by increased surface runoff \cite{lorenzoni_characterization_2015} and reduced consumption due to lower diatom abundance.
    Chlorophyll \textit{a} and functional group cell counts show positive anomalies throughout most of the first half of the time series, followed by a significant reduction after 2004, suggesting a shift. This shift occurs during the period between 2003 and 2013, when the AMO index presents mostly positive anomalies, while upwelling-related variables indicate a weak upwelling regime (low wind, high SST, deep isotherm). Although data are sparse, nutrients and biological variables show a strong increase following 2013, coinciding with a return to negative anomalies in the AMO index and more upwelling-favorable conditions.
    The annual means of diversity metrics do not follow this trend, but are reduced between 1998 and 2008. Shannon diversity specifically appears elevated past 2006 and the general pattern that emerges is that mean diversity is lower in years, where mean chlorophyll \textit{a} is high. However, in 1996 and 1997 positive anomalies in cell counts, chlorophyll \textit{a} and diversity metrics coincide.

    \begin{figure}
    \noindent\includegraphics[width=\textwidth]{fig/Figure2_ZScores_v2.pdf}
    \caption{Time series showing annual z-scores of selected environmental, climatological, and ecological variables. Red indicates negative annual anomalies and blue indicates positive annual anomalies. Deeper shades indicate larger anomalies. Chlorophyll and functional group counts were log-transformed prior to normalization. The years 1995 (n=2) and 2016 (n=5), 2017 (n=1) were removed due to insufficient monthly data to generate robust annual means. Additionally we marked years with two or three and four or more missing monthly values.}
    \label{fig:zscore}
    \end{figure}

    % Figure 3
    Figure \ref{fig:divts} shows the depth distribution and dynamics of chlorophyll \textit{a} and community diversity indices, also in relation to rainy and upwelling seasons. The bulk of the chlorophyll \textit{a} ($\sim$ 80\,\%) was measured in the upper 50 meters of the water column, with approximately 15\,\% measured between 50 and 75 meters. Chlorophyll \textit{a} exhibited marked seasonality, with the highest biomass in the upper water column during the early upwelling season (December–February, Figure \ref{fig:divts}$b$). The yearly mean chlorophyll \textit{a} biomass showed the highest variability within the upper 25 meters. In some years, particularly in 2000, 2005, 2007, 2008, and 2013, the concentration in this layer was lower than in the 25–50 meter depth range, suggesting a deeper chlorophyll maximum in those years (Figure \ref{fig:divts}$a$). The year 2013 was an outlier, particularly for the upper 25 meters, as missing monthly sampling from the upwelling season (January and February) skewed the value downwards. Nevertheless, the monthly chlorophyll \textit{a} measurements for the remainder of the year 2013 were markedly lower (65\,\% of the mean over the entire time series for these months), which recovered in 2014. Chlorophyll \textit{a} biomass below 50 meters depth showed a relatively stable distribution throughout the year and across the time series. In fall, during the rainy season, mean chlorophyll \textit{a} biomass in the upper 25 meters was markedly reduced, while the 50–75 meter depth interval showed a slight increase, indicating a shift in the chlorophyll \textit{a} maximum to lower depths under weak upwelling conditions.

    The calculated genus diversity from the microscopy cell counts generally followed the dynamics of chlorophyll \textit{a} biomass, with maximum values observed during the upwelling seasons in winter. Genus richness was consistently highest in the top \qty{25}{meters} depth interval, also showing the highest variability over the time series. Across the time series, we observed maximum richness at the start of the time series (1996-1997), with a subsequent reduction in diversity across all depths, followed by a recovery after 2011. The reduction in diversity following the start of the time series did not follow the pattern shown by chlorophyll \textit{a}, which remained relatively stable across the top \qty{100}{meters} until 2004 (see Figure \ref{fig:divts $a$}). Scaling genus richness by the cell count abundances, i.e. the Shannon index, showed a similar but less pronounced dynamic. Diversity remained high throughout 2000 and recovered more quickly but showed marked reductions in annual means, particularly in the upper 25 meters in 2006, 2009, and 2016. The seasonal signal was lost (see Figure \ref{fig:divts} $f$).

    Community evenness, expressed by the Pielou index, shows no seasonal signal and a very large variance within seasons and through depth (see Figure \ref{fig:divts} $h$). The inter-annual dynamics of evenness are remarkably stable across all depth intervals, clustered around an intermediate 0.5 value. There were marked reductions in all depths towards 2000 and for all depths above 75 meters in 2006, 2009, and 2016. Throughout 2003 to 2013 evenness was almost consistently highest in the deepest interval below 75 meters, where the community appeared to have been more evenly distributed.

@@

\subsection{Clustering analysis yields contrasting ecological regimes}
    % Figure 4
    We undertook clustering analysis on annually aggregated cell counts from microscopy and on individual monthly measurements. Figure \ref{fig:clustering}$a$ shows the results of clustering the annually aggregated communities using a linear complete linkage algorithm based on the binary Jaccard matrix. Two distinct clusters emerge from this analysis, with the first cluster encompassing the years 1996-2004 and 2014-2016, and the second cluster including the intermediate years from 2005 to 2013. 
    Inter-annual species turnover is shown in Figure \ref{fig:clustering}$b$, it reveals an increasing trend in turnover from 1996 to 2004. Turnover subsequently decreased during the period underpinning cluster 2 and increased again in 2014, marking a return to cluster 1. This pattern is also reflected in the hierarchical clustering results, Figure \ref{fig:clustering}$a$, where cluster 2 shows a lower distance in the clustering tree between the years. We additionally extracted the three dominant genera for the second cluster and the early and late parts of the first cluster, respectively. The diatom \textit{Thalassiosira} dominated the early period and was abundant in all other periods. \textit{Pseudo-nitzschia} dominated cluster 2, with the haptophyte \textit{Emiliana} being the second most abundant genus. The later part of cluster 1 is dominated by \textit{Emiliana}.

@@

    % Figure 5 - NEW
    Figure \ref{fig:clustcomp} shows density distributions of environmental variables, chlorophyll \textit{a}, and diversity indices for the two clusters. Upwelling-related variables exhibit distinct patterns between the two clusters. For example, cluster 1 experienced more frequent high wind events, a shallower \qty{21}{\celsius} isotherm, lower temperatures, reduced evaporation, and higher salinity and nutrient concentrations, particularly for nitrate. The AMO index shows a pronounced separation between the two clusters, with cluster 1 exhibiting more negative anomalies, while cluster 2 shows a higher occurrence of positive anomalies. Although there are no significant differences in genus richness and the Shannon index between clusters, there is a slightly increased median in the Pielou index for cluster 2 (Table \ref{tab:ClustCompWilcox}). The density distribution reveals some differences. Cluster 1 shows higher variance for genus richness, while cluster 2 shows greater variance and a bimodal distribution for the Shannon and Pielou indices.

@@

    The in-situ measurements that best predicted changes in the microphytoplankton community were SST, nitrate concentration, and salinity, in that order. Nitrate is a crucial nutrient for phytoplankton growth and has been identified as the growth-limiting nutrient during stratification periods in the Cariaco Basin \cite{muller-karger_scientific_2019}. For diatoms, a strong threshold response was observed above \qty{7}{\micro \molar}, indicating that intense upwelling conditions drove blooms, resulting in predictable changes in cell counts. Similarly, haptophytes and diatoms exhibited a threshold around \qtyrange{22}{24}{\celsius}, indicating thresholds at low surface temperatures driven by stronger upwelling.
